Towards a Practical Face Recognition System: \\ Robust Alignment and Illumination by Sparse Representation

### 2011-03-28 faces_system.tex
Wrote explanations for new plots and data

CMU Multi-PIE provides the most extensive test set among public
datasets. This database contains images of 337 subjects across
simultaneous variation in pose, expression, and illumination.
Of these 337 subjects, we use all of the 249 subjects present
in Session 1 as the training set. The remaining 88 subjects are
treated as ``impostors'', or invalid images. For each of the
249 training subjects, we include frontal images of 7 frontal
illuminations,\footnote{They are illuminations
$\{0,1,7,13,14,16,18\}$ of \cite{Gross2008-FGR}. For each
directional illumination, we subtract the ambient-illuminated
image 0.} taken with neutral expression. As suggested by the
work of \cite{Georghiades2001-PAMI}, these extreme frontal
illuminations would be sufficient to linearly represent other
frontal illuminations, as will also be corroborated by the next
experiment on our own dataset. For the test set, we use all 20
illuminations from Sessions 2-4, which were recorded 
over a period of several months. The dataset is
challenging due to the large number of subjects, and due to
natural variation in subject appearance over time. Table
\ref{tab:MultiPIE-recognition} shows the result of our
algorithm on each of the 3 testing sessions. Our algorithm
achieves recognition rates above $90\%$ for all three sessions.
For the test images, 
alignment was initialized automatically via the Viola and Jones'
face detector. 
To demonstrate that the sparse representation based recognition step is indeed beneficial
even when there are no impostors, we include results for recognition
based only on the alignment error residuals(i.e. S=1), shown in row 6.

\begin{table}
\caption{Recognition rates on large-scale Multi-PIE database.}
\vspace{-.2in}
\centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate & Session 2 & Session 3 & Session 4  \\
\hline
\hline
LDA$_d$ (LDA$_m$) & 5.1 (49.4)\%  & 5.9 (44.3)\% & 4.3 (47.9)\%  \\
\hline
NN$_d$ (NN$_m$)  & 26.4 (67.3)\% & 24.7 (66.2)\% & 21.9 (62.8)\%  \\
\hline
NS$_d$ (NS$_m$) &  30.8 (77.6)\% & 29.4 (74.3)\% & 24.6 (73.4)\% \\
\hline
LBP$_d$ (LBP$_m$) & 39.9 (93.3)\% & 38.1 (91.2)\% & 33.9 (92.9)\% \\
\hline
\cite{Yang2010-CVPR} & {\bf 95.2}\% & 93.4\% & 95.1\% \\
\hline
{Algorithm 1, $S=1$} & 90.7\% & 89.6\% & 87.5\% \\
\hline
{Algorithm 1} & 93.9\% & 93.8\% & 92.3\% \\
\hline
{Algorithm 1 with improved window} & 95.0\% & {\bf 96.3}\% & {\bf 97.3}\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition} }
\end{table}
\begin{figure}
\centering
{\small
\begin{tabular}{cccccc}
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/079_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/111_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/196_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/130_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/163_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/175_01_01_051_08.png} \\
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/079_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/111_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/196_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/130_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/163_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/175_02_01_051_08.png} \vspace{-.1in} \\
(a) & (b) & (c) & (d) & (e) & (f) \vspace{-.1in}
\end{tabular}
}
\caption{{\bf Representative failures from Multi-PIE}. {\bf Top:} training from Session 1; {\bf Bottom:} test images from Session 2. Due to changes in hair, glasses, beard, or pose, our alignment fails on these subjects regardless of test image illumination.}
\label{fig:failed-examples}
\vspace{-.3in}
\end{figure}

@@

\subsection{Comparison to Existing Work}
We compare our result to baseline linear-projection-based
algorithms, such as Nearest Neighbor (NN), Nearest Subspace
(NS) \cite{Lee2005-PAMI}, and Linear Discriminant Analysis
(LDA) \cite{Belhumeur1997-PAMI}.\footnote{We do not list
results on PCA \cite{Turk1991-CVPR} as its performance is
always below that of Nearest Subspace.} Since these algorithms
assume pixel-accurate alignment, they are not expected to work
well if the test is not well aligned with the training. In
Table~\ref{tab:MultiPIE-recognition}, we report the results of
those algorithms with two types of input: 1.\ the output of the
Viola and Jones' detector, indicated by a subscript ``$d$'';
2.\ the input face is aligned to the training with manually
selected outer eye corners, indicated by a subscript ``$m$''.
Notice that even with careful manual registration, these
baseline algorithms perform significantly worse than our
algorithm, which uses input directly from the face detector.
The performance drop of the LDA algorithm on Multi-PIE reported here
seems to agree with that reported already in
\cite{Gross2008-FGR}.

We also compare our result to the very recent work
\cite{Yang2010-CVPR}. Notice that in \cite{Yang2010-CVPR}, the
initial registration is again obtained from manually selected
outer eye corners. Then, a supervised hierarchical sparse
coding model based on local image descriptors is trained, which
enjoys certain translation invariant properties. With the same
training and testing sets, \cite{Yang2010-CVPR} is able to
handle any remaining misalignment and achieves state-of-the-art
performance on the CMU Multi-PIE database.
Table~\ref{tab:MultiPIE-recognition} shows that our algorithm
achieves similar or better performance on different sessions of
Multi-PIE.

Finally, we compare our result to Local Binary Patterns (LBP)
\cite{Ahonen2006-PAMI}, a local appearance descriptor which is
able to capture fine details of facial appearance and texture.
Due to its robustness to variations in illumination, facial
expression, aging and other changes, LBP has achieved the
state-of-the-art face recognition performance in the scenario
when only one sample per person is used for training
\cite{Tan06facerecognition}. In this paper, we follow the same
steps as in \cite{Ahonen2006-PAMI} to construct an LBP
descriptor for each training and testing sample. The $80\times
60$ face region is first divided into a regular $10\times 10$
grid of cells, each of size $8\times 6$ pixels. Within each
cell, the histogram of 59 uniform binary patterns is then
computed, where the patterns are generated by thresholding 8
neighboring pixels in a circle of radius 2 using the central
pixel value. Finally, the local histograms are concatenated to
produce the global descriptor vector. As suggested in
\cite{Ahonen2006-PAMI}, the recognition is performed using a
nearest neighbor classifier with Chi square distance as the
distance measure and we report the recognition rates with the
same two types of input as before. As shown in
Table~\ref{tab:MultiPIE-recognition}, LBP achieves competitive
recognition rates given manually aligned training and testing
samples, which demonstrates its robustness to illumination
changes. However, its performance is very sensitive to
misalignment.

@@

Figure \ref{fig:roc-multipie} plots the receiver operating
characteristic (ROC) curves, which are generated by sweeping
the threshold $t$ through the entire range of possible values
for each algorithm.\footnote{Rejecting invalid images not in
the entire database is much more difficult than deciding if two
face images are the same subject. Figure 8 should not be
confused with typical ROC curves for face similarity, e.g.,
\cite{PhillipsP2007}.} We can see that SCI based approach
significantly outperforms the other three algorithms. Similar
contrasts between our algorithm and baseline algorithms were
also observed for SRC in \cite{Wright2009-PAMI}, though on much
smaller datasets.

\begin{figure}
\centerline{
\includegraphics[width=3in]{figures_pami/pami_roc2.png}
}\vspace{-.2in}
\caption{{\bf ROC curves} for subject validation on Multi-PIE database.}\label{fig:roc-multipie}
\vspace{-.6in}
\end{figure}

>>>>

CMU Multi-PIE provides the most extensive test set among public datasets. This
database contains images of 337 subjects across simultaneous variation in pose,
expression, and illumination.  Of these 337 subjects, we use all of the 249
subjects present in Session 1 as the training set. The remaining 88 subjects
are treated as ``impostors'', or invalid images. For each of the 249 training
subjects, we include frontal images of 7 frontal illuminations,\footnote{They
are illuminations $\{0,1,7,13,14,16,18\}$ of \cite{Gross2008-FGR}. For each
directional illumination, we subtract the ambient-illuminated image 0.} taken
with neutral expression. As suggested by the work of
\cite{Georghiades2001-PAMI}, these extreme frontal illuminations would be
sufficient to linearly represent other frontal illuminations, as will also be
corroborated by the next experiment on our own dataset. For the test set, we
use all 20 illuminations from Sessions 2-4, which were recorded over a period
of several months. The dataset is challenging due to the large number of
subjects, and due to natural variation in subject appearance over time.  Line 5
of Table \ref{tab:MultiPIE-recognition} shows the result of our recognition
pipeline on each of the 3 testing sessions. Our algorithm achieves recognition
rates above $90\%$ for all three sessions.  For the test images, our iterative
alignment was initialized automatically via the Viola and Jones' face detector.

\begin{table}
\caption{Recognition rates on the Multi-PIE database for different pairings of alignment
and recognition stages.}
\vspace{-.2in}
\centerline{
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c| }
\hline
\backslashbox{Rec.}{Align.} 
& \multicolumn{3}{|c|}{Face Detector} 
& \multicolumn{3}{|c|}{Manual} 
& \multicolumn{3}{|c|}{Iterative} 
\\
\hline
Session $\rightarrow$	& 2		&3			&4			& 2		&3			&4			& 2		&3			&4		\\
\hline
NS	& 30.8\%	& 29.4\%	& 24.6\%	& 77.6\%	& 74.3\%	& 73.4\%	& 84.5\%	& 82.3\%	& 81.4\% \\
\hline
NN	& 26.4\%	& 24.7\%	& 21.9\%	& 67.3\%	& 66.2\%	& 62.8\%	& 73.5\%	& 69.6\%	& 69.3\% \\
\hline
LDA	& 5.1\%		& 5.9\%		& 4.3\%		& 49.4\%	& 44.3\%	& 47.9\%	& 91.0\%	& 89.9\%	& 88.1\% \\
\hline
LBP	& 39.9\%	& 38.1\%	& 33.9\%	& 93.3\%	& 91.2\%	& 92.9\%	& {\bf 95.2\%}	& {\bf 94.7\%}	& {\bf 93.5\%} \\
\hline
SRC	&			&			&			&			&			&			& 93.9\%	& 93.8\%	& 92.3\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition} }
\end{table}

\begin{figure}
\centering
{\small
\begin{tabular}{cccccc}
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/079_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/111_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/196_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/130_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/163_01_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/175_01_01_051_08.png} \\
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/079_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/111_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/196_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/130_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/163_02_01_051_08.png}  &
\includegraphics[width=0.9in,clip=true]{figures_pami/multipie_failed/175_02_01_051_08.png} \vspace{-.1in} \\
(a) & (b) & (c) & (d) & (e) & (f) \vspace{-.1in}
\end{tabular}
}
\caption{{\bf Representative failures from Multi-PIE}. {\bf Top:} training from Session 1; {\bf Bottom:} test images from Session 2. Due to changes in hair, glasses, beard, or pose, our alignment fails on these subjects regardless of test image illumination.}
\label{fig:failed-examples}
\vspace{-.3in}
\end{figure}

@@

\subsection{Comparison to Existing Work}
% Classical algorithms
We compare our result to baseline linear-projection-based
algorithms, such as Nearest Neighbor (NN), Nearest Subspace
(NS) \cite{Lee2005-PAMI}, and Linear Discriminant Analysis
(LDA) \cite{Belhumeur1997-PAMI}.\footnote{We do not list
results on PCA \cite{Turk1991-CVPR} as its performance is
always below that of Nearest Subspace.} Since these algorithms
assume pixel-accurate alignment, they are not expected to work
well if the test image is not well aligned with the training. In
Table~\ref{tab:MultiPIE-recognition}, we report the results of
these classical algorithms with three types of testing image alignment: 
1.\ alignment from the Viola and Jones' detector,
2.\ alignment via manually selected outer eye corners, and 
3.\ the output of our iterative alignment algorithm.
The performance drop of the LDA algorithm on Multi-PIE reported here
seems to agree with that reported already in
\cite{Gross2008-FGR}.  All of the classical algorithms benefit
greatly from being paired with our iterative alignment algorithm.

% LBP
Finally, we compare our result to Local Binary Patterns (LBP)
\cite{Ahonen2006-PAMI}, a local appearance descriptor which is able to capture
fine details of facial appearance and texture.  Due to its robustness to
variations in illumination, facial expression, aging and other changes, LBP has
achieved the state-of-the-art face recognition performance in the scenario when
only one sample per person is used for training \cite{Tan06facerecognition}. In
this paper, we follow the same steps as in \cite{Ahonen2006-PAMI} to construct
an LBP descriptor for each training and testing sample. The $80\times 60$ face
region is first divided into a regular $10\times 10$ grid of cells, each of
size $8\times 6$ pixels. Within each cell, the histogram of 59 uniform binary
patterns is then computed, where the patterns are generated by thresholding 8
neighboring pixels in a circle of radius 2 using the central pixel value.
Finally, the local histograms are concatenated to produce the global descriptor
vector. As suggested in \cite{Ahonen2006-PAMI}, the recognition is performed
using a nearest neighbor classifier with Chi square distance as the distance
measure and we report the recognition rates with the same two types of input as
before.  Recognition results for LBP on Multi-PIE are shown in
Table~\ref{tab:MultiPIE-recognition}.  We attribute the improvement in
performance of LBP over SRC in this experiment to its robustness to
illumination components that are not in the training set.  Some of the test
images used in this experiment are taken under illuminations not present in the
training subset.  Note that, like the other classical algorithms, the
performance of LBP is still sensitive to misalignment. This is notable given
that LBP is often applied without any special alignment.

\begin{table}
\caption{Recognition rates on the Multi-PIE database for \cite{Yang2010-CVPR} and variants
of Algorithm 1}
\vspace{-.2in}
\centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate & Session 2 & Session 3 & Session 4  \\
\hline
\cite{Yang2010-CVPR} & {\bf 95.2}\% & 93.4\% & 95.1\% \\
\hline
{Algorithm 1, $S=1$} & 90.7\% & 89.6\% & 87.5\% \\
\hline
{Algorithm 1} & 93.9\% & 93.8\% & 92.3\% \\
\hline
{Algorithm 1 with improved window} & 95.0\% & {\bf 96.3}\% & {\bf 97.3}\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition2} }
\vspace{-.3in}
\end{table}

To demonstrate that the sparse representation based recognition step is indeed
beneficial even when there are no impostors, we include results for recognition
based only on the alignment error residuals(i.e. $S=1$), shown in 
Table \ref{tab:MultiPIE-recognition2}, row 6.

We also compare our result to the recent work
\cite{Yang2010-CVPR}. Notice that in \cite{Yang2010-CVPR}, the
initial registration is again obtained from manually selected
outer eye corners. Then, a supervised hierarchical sparse
coding model based on local image descriptors is trained, which
enjoys certain translation invariant properties. With the same
training and testing sets, \cite{Yang2010-CVPR} is able to
handle any remaining misalignment and achieves state-of-the-art
performance on the CMU Multi-PIE database.
Table~\ref{tab:MultiPIE-recognition2} shows that our algorithm
achieves similar or better performance on different sessions of
Multi-PIE.

@@

Figure \ref{fig:roc-multipie} plots the receiver operating characteristic (ROC)
curves, which are generated by sweeping the threshold $t$ through the entire
range of possible values for each algorithm.\footnote{Rejecting invalid images
not in the entire database is much more difficult than deciding if two face
images are the same subject. Figure 8 should not be confused with typical ROC
curves for face similarity, e.g., \cite{PhillipsP2007}.} On the left we can see
that the SCI based recognition approach significantly outperforms the other
algorithms, including LBP, even when all algorithms are coupled with our
proposed iterative alignment.  In the right plot we again see that classical
algorithms, and even LBP, are very sensitive to alignment.  Similar contrasts
between our algorithm and baseline algorithms were also observed for SRC in
\cite{Wright2009-PAMI}, though on much smaller datasets.
\begin{figure}
\centerline{
	\begin{tabular}{cc}
	\includegraphics[width=3in]{figures_pami/pami_roc_revision2.png} &
	\includegraphics[width=3in]{figures_pami/pami_roc2.png} \\
	{\small With iterative alignment} & {\small Classical Algorithms with manual alignment} 
\\
	\end{tabular}
}
\caption{{\bf ROC curves} for subject validation on Multi-PIE database, (a) with iterative alignment,
and (b) for the classical algorithms with manual alignment}\label{fig:roc-multipie}
%\vspace{-.6in}
\end{figure}
